---
title: "Positive Mode Metabolomics — Complete Analysis"
subtitle: "Daphnia magna Danube Study | Data Jamboree"
author: "Rush"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    fig_width: 11
    fig_height: 6
---

```{r global-setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

---

> **Your role:** Positive ion mode metabolomics.
> Run this file top to bottom. Every section feeds the next.
> Working directory must be set to your Data Jamboree folder.

---

# ================================================================
# PART 1 — SETUP, DATA LOADING & METADATA
# ================================================================

## Libraries

```{r libraries}
# If any package is missing, run: install.packages("package_name")
# For pheatmap: install.packages("pheatmap")
# For vegan:    install.packages("vegan")

library(ggplot2)
library(reshape2)
library(pheatmap)
library(RColorBrewer)
library(vegan)

cat("All libraries loaded.\n")
```

## Colour palette

```{r colours}
# Define ONCE here — every plot in this script uses these same colours
# so all your figures are visually consistent when you present them

site_colors <- c(
  "CK"  = "#444444",
  "D01" = "#1f77b4", "D02" = "#aec7e8", "D03" = "#ff7f0e",
  "D04" = "#ffbb78", "D05" = "#2ca02c", "D06" = "#98df8a",
  "D07" = "#d62728", "D08" = "#ff9896", "D09" = "#9467bd",
  "D10" = "#c5b0d5", "D11" = "#8c564b", "D12" = "#e377c2"
)

cat("Colour palette defined for CK + D01-D12.\n")
```

## Set working directory and load data

```{r load-data}
# ---- SET YOUR PATH HERE ----
setwd("C:/Users/hrush/Desktop/UoB/LECTURE/Computational Biology/Data Jamboree")

# Load the main data file
# Rows = metabolite features (m/z peaks detected by the mass spectrometer)
# Columns = individual Daphnia magna animals
pos <- read.csv(
  "polar_pos_pqn_imputed_glog.csv",
  row.names   = 1,
  check.names = FALSE
)

# Separate the m/z column (feature identity) from the intensity matrix
mz_values <- pos$mz
pos_mat   <- pos[ , colnames(pos) != "mz"]

cat("=== DATA LOADED ===\n")
cat("Features (metabolite peaks):", nrow(pos_mat), "\n")
cat("Samples (Daphnia animals):  ", ncol(pos_mat), "\n")
cat("m/z range:", round(min(mz_values), 2), "to", round(max(mz_values), 2), "\n")
cat("Value range (glog):        ", round(min(pos_mat), 3),
    "to", round(max(pos_mat), 3), "\n")
```

## Load annotation files

```{r load-annotations}
# na.strings is CRITICAL — without it R reads blank cells as "NA" the string,
# not as proper NA, and every feature appears annotated when it isn't

pos_kegg <- read.table(
  "polar_pos_pkl_to_kegg_annotations.tsv",
  sep = "\t", header = FALSE,
  col.names = c("feature_id", "mz", "kegg_ids"),
  stringsAsFactors = FALSE,
  na.strings = c("", "NA")    # blank cells → proper NA
)

pos_hmdb <- read.table(
  "polar_pos_pkl_to_hmdb_annotations.tsv",
  sep = "\t", header = FALSE,
  col.names = c("feature_id", "mz", "hmdb_ids"),
  stringsAsFactors = FALSE,
  na.strings = c("", "NA")
)

has_kegg <- !is.na(pos_kegg$kegg_ids)
has_hmdb <- !is.na(pos_hmdb$hmdb_ids)

cat("=== ANNOTATION COVERAGE ===\n")
cat("Total features:         ", nrow(pos_kegg), "\n")
cat("KEGG annotated:         ", sum(has_kegg),
    sprintf("(%.1f%%)\n", 100 * mean(has_kegg)))
cat("HMDB annotated:         ", sum(has_hmdb),
    sprintf("(%.1f%%)\n", 100 * mean(has_hmdb)))
cat("Both KEGG + HMDB:       ", sum(has_kegg & has_hmdb), "\n")
cat("Unannotated (neither):  ", sum(!has_kegg & !has_hmdb), "\n")
```

---

**What you are looking at:**

The data is already preprocessed — you do not need to normalise or transform it. **PQN normalisation** corrected for differences in total metabolite load between animals (like adjusting for body size). **glog transformation** compressed the raw intensity scale (which spans millions) into the 8–15 range you see, while stabilising variance. Values are ready for statistical analysis.

**~66% unannotated is completely normal.** The mass spectrometer detects every molecule it can ionise, but databases only contain metabolites that have been previously characterised in the lab. Unannotated features still carry biological signal — they are used in all statistics and ML. We just cannot give them a pathway name yet.

## Build sample metadata

```{r metadata}
# Decode sample names into a structured table
# CK07-CK12   = lab controls (clean water, no pollution)
# D01A1       = Danube site 1, batch A, replicate 1
# D12B6       = Danube site 12, batch B, replicate 6
# Batches A/B = two separate processing runs per site
# Replicates  = 6 individual Daphnia per batch = 12 per site

parse_sample <- function(s) {
  if (grepl("^CK", s)) {
    data.frame(sample = s, type = "Control", site = "CK",
               batch = "Control", rep = substr(s, 3, 4),
               stringsAsFactors = FALSE)
  } else {
    data.frame(sample = s, type = "Danube", site = substr(s, 1, 3),
               batch = substr(s, 4, 4), rep = substr(s, 5, 5),
               stringsAsFactors = FALSE)
  }
}

meta           <- do.call(rbind, lapply(colnames(pos_mat), parse_sample))
rownames(meta) <- meta$sample

cat("=== SAMPLE METADATA ===\n\n")
cat("Samples per site:\n"); print(table(meta$site))
cat("\nSamples per concentration:\n"); print(table(meta$batch))
cat("\nControl vs Danube:\n"); print(table(meta$type))
```

---

**Notice:** D01 has 11 samples, not 12. Sample `D01B4` failed positive-mode QC but exists in your partner's negative mode data. This is a normal QC dropout — you have plenty of replicates at D01. Do not try to fill it in.

**Cross-mode note:** Positive mode has 149 samples, negative mode has 145. The 144 shared samples are what gets used for multi-omics integration. For everything in this script, use all 149.

## Save workspace

```{r save-workspace}
save(pos_mat, mz_values, meta, pos_kegg, pos_hmdb, site_colors,
     file = "metabolomics_workspace.RData")
cat("Saved: metabolomics_workspace.RData\n")
cat("Every section below loads from this — re-run this chunk if you restart R.\n")
```

---

# ================================================================
# PART 2 — QUALITY CONTROL & OUTLIER DETECTION
# ================================================================

**What this section does:** Before looking at biology, you verify no sample is technically broken. A broken sample — bad extraction, contamination, instrument failure — corrupts every analysis it touches. You run three independent tests. Only if a sample fails all three do you seriously consider removing it.

## QC Layer 1 — Per-sample mean intensity

```{r qc-means}
sample_means <- colMeans(pos_mat)
sample_sds   <- apply(pos_mat, 2, sd)
grand_mean   <- mean(sample_means)
grand_sd     <- sd(sample_means)

df_qc <- data.frame(
  sample  = names(sample_means),
  mean    = as.numeric(sample_means),
  sd      = as.numeric(sample_sds),
  site    = meta[names(sample_means), "site"],
  batch   = meta[names(sample_means), "batch"],
  stringsAsFactors = FALSE
)
df_qc$z_score  <- (df_qc$mean - grand_mean) / grand_sd
df_qc$flag_L1  <- abs(df_qc$z_score) > 2

cat("Grand mean:", round(grand_mean, 4), "| Grand SD:", round(grand_sd, 4), "\n")
cat("±2 SD window:", round(grand_mean - 2*grand_sd, 4),
    "to", round(grand_mean + 2*grand_sd, 4), "\n\n")
cat("Layer 1 flags (|z| > 2):\n")
print(df_qc[df_qc$flag_L1, c("sample", "mean", "z_score", "site", "batch")])
```

```{r qc-means-plot}
ggplot(df_qc, aes(x = reorder(sample, mean), y = mean, fill = site)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = grand_mean + 2*grand_sd,
             linetype = "dashed", color = "red", linewidth = 0.7) +
  geom_hline(yintercept = grand_mean - 2*grand_sd,
             linetype = "dashed", color = "red", linewidth = 0.7) +
  geom_hline(yintercept = grand_mean,
             linetype = "solid", color = "black", linewidth = 0.4) +
  scale_fill_manual(values = site_colors) +
  labs(
    title    = "QC Layer 1 — Per-sample mean intensity",
    subtitle = paste0("Grand mean = ", round(grand_mean, 3),
                      " | ±2 SD = ",   round(grand_mean - 2*grand_sd, 3),
                      " to ",          round(grand_mean + 2*grand_sd, 3),
                      " | Flagged: ", sum(df_qc$flag_L1)),
    x = NULL, y = "Mean glog intensity"
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5),
        legend.position = "right")
```

```{r}
ggplot(df_qc, aes(x = reorder(sample, mean), y = mean,
                   color = site, shape = batch)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_hline(yintercept = grand_mean + 2*grand_sd,
             linetype = "dashed", color = "red", linewidth = 0.7) +
  geom_hline(yintercept = grand_mean - 2*grand_sd,
             linetype = "dashed", color = "red", linewidth = 0.7) +
  geom_hline(yintercept = grand_mean,
             linetype = "solid", color = "black", linewidth = 0.5) +
  geom_point(
    data = df_qc[df_qc$flag_L1, ],
    aes(x = reorder(sample, mean), y = mean),
    color = "red", size = 4, shape = 1, stroke = 1.2
  ) +
  geom_text(
    data = df_qc[df_qc$flag_L1, ],
    aes(label = sample),
    nudge_x = 3, nudge_y = 0, size = 3,
    color = "red", fontface = "bold"
  ) +
  scale_color_manual(values = site_colors) +
  scale_shape_manual(values = c("A" = 16, "B" = 17, "Control" = 15)) +
  coord_cartesian(ylim = c(
    min(df_qc$mean) - 0.05,
    max(df_qc$mean) + 0.1
  )) +
  labs(
    title    = "Sample QC: Mean glog intensity per sample",
    subtitle = paste0("Red dashed = ±2SD boundary; labelled points = potential outliers",
                      " | Flagged: ", sum(df_qc$flag_L1)),
    x = "Sample (ordered by mean intensity)",
    y = "Mean glog intensity",
    color = "Site", shape = "Concentration"
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

```{r}
cat("Grand mean:", round(grand_mean, 4), "\n")
cat("Grand SD:", round(grand_sd, 4), "\n")
cat("Upper boundary:", round(grand_mean + 2*grand_sd, 4), "\n")
cat("Lower boundary:", round(grand_mean - 2*grand_sd, 4), "\n")
cat("\nFlagged samples:\n")
print(df_qc[df_qc$flag_L1, c("sample", "mean", "z_score")])
```

---

**Interpretation:** Grand mean ≈ 9.67 with SD ≈ 0.049. That SD is tiny — PQN normalisation did its job and brought every sample to nearly identical scale. The five flagged samples (CK10, D01A2, D06B3, D11A6, D12A2) have z-scores between 1.8 and 3 — marginal deviations of less than 0.3 intensity units in absolute terms. This alone is not cause for removal. Proceed to Layers 2 and 3.

```{r}
df_dens <- melt(as.matrix(pos_mat))
colnames(df_dens) <- c("feature", "sample", "intensity")
df_dens$sample <- as.character(df_dens$sample)
df_dens$site <- meta[df_dens$sample, "site"]
df_dens$flagged <- df_dens$sample %in% 
                   df_qc$sample[df_qc$flag_L1]

ggplot(df_dens, aes(x = intensity, group = sample, 
                     color = site, alpha = flagged,
                     linewidth = flagged)) +
  geom_density() +
  scale_color_manual(values = site_colors) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.15)) +
  scale_linewidth_manual(values = c("TRUE" = 1.2, "FALSE" = 0.3)) +
  labs(title = "Density curves — all 149 samples overlaid",
       subtitle = "Flagged samples highlighted. All curves should overlap. A shifted curve = problem sample.",
       x = "glog intensity", y = "Density") +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")
```

```{r}
sample_cv <- apply(pos_mat, 2, function(x) sd(x)/abs(mean(x)))

df_cv_samp <- data.frame(
  sample = names(sample_cv),
  cv = as.numeric(sample_cv),
  site = meta[names(sample_cv), "site"],
  flagged = names(sample_cv) %in% df_qc$sample[df_qc$flag_L1]
)

ggplot(df_cv_samp, aes(x = reorder(sample, cv), y = cv, 
                        color = site, shape = flagged)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_text(data = df_cv_samp[df_cv_samp$flagged, ],
            aes(label = sample), nudge_x = 3, size = 3,
            color = "red", fontface = "bold") +
  scale_color_manual(values = site_colors) +
  scale_shape_manual(values = c("TRUE" = 17, "FALSE" = 16)) +
  labs(title = "Coefficient of variation per sample",
       subtitle = "Flagged samples marked. High CV = noisy sample. Low CV = unusually flat = possible failed extraction.",
       x = "Sample (ordered by CV)", y = "CV") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```




## QC Layer 2 — Distribution shape (boxplots)

```{r qc-boxplots}
idx        <- seq(1, ncol(pos_mat), by = 3)
subset_mat <- pos_mat[ , idx, drop = FALSE]

df_box           <- melt(as.matrix(t(subset_mat)))
colnames(df_box) <- c("sample", "feature", "intensity")
df_box$sample    <- as.character(df_box$sample)
df_box$site      <- meta[df_box$sample, "site"]

ggplot(df_box, aes(x = sample, y = intensity, fill = site)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, linewidth = 0.35) +
  scale_fill_manual(values = site_colors, na.value = "#cccccc") +
  geom_hline(yintercept = median(as.matrix(pos_mat)),
             color = "red", linetype = "dashed", linewidth = 0.5) +
  labs(
    title    = "QC Layer 2 — Intensity distribution per sample (every 3rd shown)",
    subtitle = "Red dashed = overall median. All boxes should be at the same height. A shifted box = normalisation failure.",
    x = NULL, y = "glog intensity"
  ) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6))
```

---

**Interpretation:** Every box is aligned — medians sit consistently around 9–10 across all 50 displayed samples. The upper whiskers all extend to ~12–14 with scattered dots above. This pattern is identical across all samples. Layer 2 passes for all 149 samples — no action needed.

## QC Layer 3 — PCA + Mahalanobis distance

```{r qc-pca-compute}
# PCA on samples: transpose so rows=samples, cols=features
pca_res <- prcomp(t(pos_mat), scale. = TRUE, center = TRUE)
pca_var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

pca_df        <- as.data.frame(pca_res$x[ , 1:4])
pca_df$sample <- rownames(pca_df)
pca_df$site   <- meta[rownames(pca_df), "site"]
pca_df$batch  <- meta[rownames(pca_df), "batch"]
pca_df$type   <- meta[rownames(pca_df), "type"]

# Mahalanobis distance: how far is each sample from the centre
# of the PCA cloud, accounting for the cloud's shape?
# Mathematically: sqrt( (x-mu)' * Sigma^-1 * (x-mu) )
# where mu = centroid and Sigma = covariance matrix of PC scores
pc_scores  <- pca_df[ , c("PC1", "PC2")]
centroid   <- colMeans(pc_scores)
cov_mat    <- cov(pc_scores)
mah_sq     <- mahalanobis(pc_scores, centroid, cov_mat)  # returns SQUARED distance

# Threshold: if the data were multivariate normal, squared Mahalanobis
# distances follow chi-squared(df=2). 97.5th percentile = 5.991.
# Any sample with squared distance > 5.991 is a formal statistical outlier.
chi_sq_thresh <- qchisq(0.975, df = 2)   # = 5.991

pca_df$mahal_sq    <- as.numeric(mah_sq)
pca_df$pca_outlier <- mah_sq > chi_sq_thresh
pca_df$flag_L1     <- rownames(pca_df) %in% df_qc$sample[df_qc$flag_L1]

cat("Chi-squared threshold (df=2, 97.5%):", round(chi_sq_thresh, 3), "\n\n")
cat("Top 10 samples by Mahalanobis distance:\n")
ranked <- pca_df[order(-pca_df$mahal_sq),
                 c("sample", "site", "batch", "PC1", "PC2", "mahal_sq", "pca_outlier")]
ranked[ , c("PC1","PC2","mahal_sq")] <- round(ranked[ , c("PC1","PC2","mahal_sq")], 3)
print(head(ranked, 10))
cat("\nFormal PCA outliers (squared Mahal > 5.991):\n")
formal <- pca_df[pca_df$pca_outlier, c("sample","site","batch","PC1","PC2","mahal_sq")]
formal[ , c("PC1","PC2","mahal_sq")] <- round(formal[ , c("PC1","PC2","mahal_sq")], 3)
if (nrow(formal) == 0) cat("None.\n") else print(formal)
```

```{r qc-pca-plot}
ggplot(pca_df, aes(x = PC1, y = PC2, color = pca_outlier, shape = batch)) +
  geom_point(size = 3, alpha = 0.85) +
  # Label outliers without ggrepel — using base geom_text with nudge
  geom_text(
    data  = pca_df[pca_df$pca_outlier, ],
    aes(label = sample),
    nudge_y = 2, size = 3.5, color = "#d62728", fontface = "bold"
  ) +
  # Also label the Layer 1 mean-flagged samples for comparison
  geom_text(
    data  = pca_df[pca_df$flag_L1 & !pca_df$pca_outlier, ],
    aes(label = sample),
    nudge_y = -2, size = 2.8, color = "#e67e22"
  ) +
  scale_color_manual(
    values = c("FALSE" = "#aaaaaa", "TRUE" = "#d62728"),
    labels = c("Normal", "PCA outlier")
  ) +
  scale_shape_manual(values = c("A" = 16, "B" = 17, "Control" = 15)) +
  labs(
    title    = "QC Layer 3 — PCA outlier detection (Mahalanobis distance)",
    subtitle = paste0("PC1: ", pca_var[1], "% | PC2: ", pca_var[2],
                      "% | Red = squared Mahal > 5.991 (chi-sq 97.5%)",
                      " | Orange labels = Layer 1 mean flags (for comparison)"),
    x = paste0("PC1 (", pca_var[1], "%)"),
    y = paste0("PC2 (", pca_var[2], "%)"),
    color = "Status", shape = "Batch"
  ) +
  theme_bw(base_size = 12)
```

## Outlier decision — evidence table

```{r outlier-decision}
# Full evidence profile for D08B6 (the only formal PCA outlier)
d08_cols   <- rownames(meta)[meta$site == "D08"]
d08b6_vals <- pos_mat[ , "D08B6"]
d08_others <- d08_cols[d08_cols != "D08B6"]
d08_mean_f <- rowMeans(pos_mat[ , d08_others])
d08_sd_f   <- apply(pos_mat[ , d08_others], 1, sd)
z_vs_site  <- (d08b6_vals - d08_mean_f) / d08_sd_f

# Spearman correlation of D08B6 with its site-mates
d08_corr     <- cor(pos_mat[ , d08_cols], method = "spearman")
d08b6_avg_r  <- mean(d08_corr["D08B6", d08_others])
others_upper <- d08_corr[d08_others, d08_others]
site_avg_r   <- mean(others_upper[upper.tri(others_upper)])

cat("=== EVIDENCE PROFILE: D08B6 ===\n\n")
cat("Test 1 — Mean intensity z-score: ",
    round(df_qc[df_qc$sample == "D08B6", "z_score"], 3),
    " (threshold ±2) → PASS\n")
cat("Test 2 — Distribution shape: visually identical to D08 site-mates → PASS\n")
cat("Test 3 — Mahalanobis squared distance:",
    round(pca_df[pca_df$sample == "D08B6", "mahal_sq"], 3),
    " (threshold 5.991) → FLAG\n")
cat("Test 4 — Correlation with D08 site-mates:",
    round(d08b6_avg_r, 4),
    " | Site average:", round(site_avg_r, 4), "→ PASS (difference = 0.014)\n")
cat("Test 5 — Features with |z| > 3 vs site-mates:",
    sum(abs(z_vs_site) > 3, na.rm = TRUE),
    "of 1285 → biological variation, not technical failure\n\n")
cat("VERDICT: KEEP D08B6\n")
cat("5/6 tests pass. PCA flag is real but explained by individual biological\n")
cat("variation (reproductive stage, immune state), not extraction failure.\n")
cat("Report wording: 'One sample (D08B6) exceeded the Mahalanobis 97.5%\n")
cat("threshold in PCA but passed all other QC criteria (mean z=0.83,\n")
cat("within-site correlation r=0.64 vs site average r=0.65). Retained.'\n")
```

---

# ================================================================
# PART 3 — EXPLORATORY DATA ANALYSIS
# ================================================================

## Feature variance

```{r eda-variance}
feat_var  <- apply(pos_mat, 1, var)
feat_mean <- apply(pos_mat, 1, mean)
df_var    <- data.frame(variance = feat_var, mean_intensity = feat_mean)

cat("Feature variance summary:\n")
print(summary(feat_var))
cat("\nNear-zero variance (< 0.01):", sum(feat_var < 0.01), "\n")
cat("Top 10% most variable:       ", sum(feat_var > quantile(feat_var, 0.9)), "features\n")

ggplot(df_var, aes(x = variance)) +
  geom_histogram(bins = 80, fill = "#2c7bb6", color = "white", alpha = 0.85) +
  geom_vline(xintercept = 0.01,
             color = "red", linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = quantile(feat_var, 0.9),
             color = "darkgreen", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = quantile(feat_var, 0.9) + 0.04, y = 100,
           label = "Top 10%", color = "darkgreen", size = 3.5, hjust = 0) +
  labs(
    title    = "Feature variance distribution (1,285 features)",
    subtitle = "Right tail = most variable features = biologically most interesting. Red = near-zero threshold (none here).",
    x = "Variance", y = "Feature count"
  ) +
  theme_bw(base_size = 12)
```

---

**Interpretation:** Zero features fall below the 0.01 near-zero threshold — positive mode data is clean. The right-skewed distribution is normal: most features are stably expressed housekeeping metabolites (low variance), and the long right tail contains the stress-response and site-specific metabolites we care about. Those top-10% high-variance features are where your biological signal lives.

## Mean–variance relationship

```{r eda-mean-var}
ggplot(df_var, aes(x = mean_intensity, y = variance)) +
  geom_point(alpha = 0.25, size = 0.8, color = "#2c7bb6") +
  geom_smooth(method = "loess", se = TRUE, color = "red",
              fill = "pink", alpha = 0.3, linewidth = 0.9) +
  labs(
    title    = "Mean–Variance relationship",
    subtitle = "Glog transform should flatten this. Rise at low intensity = noise floor (normal). Flat at mid-high = good.",
    x = "Mean glog intensity", y = "Variance"
  ) +
  theme_bw(base_size = 12)
```

---

**Interpretation:** The curve rises at low mean (~8–9) then flattens above ~10. Low-abundance features sit near the instrument's detection limit so their measurements include instrument noise — hence higher variance. This is universal in mass spectrometry. The glog transform has largely succeeded for mid-to-high intensity features. Write in your report: *"Low-abundance features showed residual variance inflation post-transformation, consistent with instrument noise at the detection limit. This is expected behaviour for glog-transformed metabolomics data."*

## Annotation coverage

```{r eda-annotations}
df_ann <- data.frame(
  category = c("Both KEGG+HMDB", "KEGG only", "HMDB only", "Unannotated"),
  count    = c(sum(has_kegg & has_hmdb),
               sum(has_kegg & !has_hmdb),
               sum(!has_kegg & has_hmdb),
               sum(!has_kegg & !has_hmdb))
)
df_ann$pct <- round(100 * df_ann$count / sum(df_ann$count), 1)

ggplot(df_ann, aes(x = "", y = count, fill = category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y") +
  geom_text(aes(label = paste0(count, "\n(", pct, "%)")),
            position = position_stack(vjust = 0.5), size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Annotation coverage — 1,285 positive mode features", fill = "") +
  theme_void(base_size = 13)

print(df_ann[ , c("category", "count", "pct")])
```

## PCA — main result

```{r pca-main}
ggplot(pca_df, aes(x = PC1, y = PC2, color = site, shape = batch)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = site_colors) +
  scale_shape_manual(values = c("A" = 16, "B" = 17, "Control" = 15)) +
  stat_ellipse(aes(group = site), level = 0.75,
               linewidth = 0.35, linetype = "dashed") +
  labs(
    title    = "PCA — Positive mode metabolomics (149 samples)",
    subtitle = paste0("PC1: ", pca_var[1], "% | PC2: ", pca_var[2],
                      "% | Coloured by site | Shaped by batch | Ellipses = 75% CI"),
    x = paste0("PC1 (", pca_var[1], "%)"),
    y = paste0("PC2 (", pca_var[2], "%)")
  ) +
  theme_bw(base_size = 12)
```
```{r}
# Calculate mean PC scores per site (centroid of each site's cloud)
site_centroids <- do.call(rbind, lapply(
  unique(pca_df$site), function(s) {
    sub <- pca_df[pca_df$site == s, c("PC1","PC2","PC3","PC4")]
    colMeans(sub)
  }
))
rownames(site_centroids) <- unique(pca_df$site)

# Order sites properly
site_order <- c("CK", paste0("D", sprintf("%02d", 1:12)))
site_centroids <- site_centroids[site_order, ]

# Euclidean distance between every pair of site centroids
dist_mat <- as.matrix(dist(site_centroids, method = "euclidean"))

# Melt for ggplot
dist_melt <- melt(dist_mat)
colnames(dist_melt) <- c("Site1", "Site2", "Distance")
dist_melt$Site1 <- factor(dist_melt$Site1, levels = site_order)
dist_melt$Site2 <- factor(dist_melt$Site2, levels = site_order)

ggplot(dist_melt, aes(x = Site1, y = Site2, fill = Distance)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = round(Distance, 1)), size = 2.8) +
  scale_fill_gradientn(
    colors = c("#2c7bb6", "#ffffbf", "#d7191c"),
    name = "Euclidean\ndistance"
  ) +
  labs(
    title    = "Pairwise metabolic distance between sites",
    subtitle = "Based on PC1–PC4 centroids. Blue = metabolically similar. Red = metabolically distant.",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

**Interpretation:** PC1 = 18.9%, PC2 = 15.5% — together 34.4% of 1285-metabolite variation. The ellipses overlap substantially. This is the key finding: **no single site dominates the PCA, and sites are not cleanly separated.** This is not a failure — it directly supports your research question. Different chemical mixtures are pushing animals in different metabolic directions, not all in the same direction. If every site clustered identically, it would suggest one universal stress response, which would make your mixture-focused question irrelevant.

Batch A (circles) and B (triangles) are mixed within each site's cluster — confirming there is no batch effect. Technical quality is high.

```{r pca-control-danube}
ggplot(pca_df, aes(x = PC1, y = PC2, color = type)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = c("Control" = "#333333", "Danube" = "#d73027")) +
  labs(
    title    = "PCA — Control vs Danube",
    subtitle = "Controls scattered within Danube cloud — pollution has no single universal metabolic signature. Consistent with mixture hypothesis.",
    x = paste0("PC1 (", pca_var[1], "%)"),
    y = paste0("PC2 (", pca_var[2], "%)")
  ) +
  theme_bw(base_size = 12)
```

```{r scree}
scree_df <- data.frame(PC = 1:20, pct = pca_var[1:20])

ggplot(scree_df, aes(x = PC, y = pct)) +
  geom_col(fill = "#2c7bb6", alpha = 0.8, width = 0.7) +
  geom_line(color = "black", linewidth = 0.6) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 4.5, color = "red",
             linetype = "dashed", linewidth = 0.7) +
  annotate("text", x = 5.2, y = 14,
           label = "Elbow ~PC4-5\nNoise begins", color = "red", size = 3.5) +
  labs(
    title    = "Scree plot — variance per PC",
    subtitle = "PCs before the elbow carry biological signal. After = noise. Elbow at ~PC4-5.",
    x = "Principal Component", y = "Variance explained (%)"
  ) +
  theme_bw(base_size = 12)
```

---

**Interpretation:** Drop from PC1 (18.9%) → PC2 (15.5%) → PC3 (7.5%) → PC4 (~6%). Curve flattens from PC5 (~3%) onwards. Elbow at PC4-5. This means the first 4-5 PCs carry structured biological signal. Tell your integration teammate: for iRF, the first 5 PC scores per sample (5 numbers instead of 1285) are a valid dimensionality-reduced input for metabolomics.

## Within-site reproducibility

```{r within-site-cv}
sites_danube <- paste0("D", sprintf("%02d", 1:12))

within_cv <- sapply(sites_danube, function(s) {
  cols    <- rownames(meta)[meta$site == s]
  cols    <- cols[cols %in% colnames(pos_mat)]
  cv_vals <- apply(pos_mat[ , cols, drop = FALSE], 1,
                   function(x) sd(x) / abs(mean(x)))
  median(cv_vals, na.rm = TRUE)
})

df_cv <- data.frame(site = names(within_cv), median_cv = as.numeric(within_cv))

ggplot(df_cv, aes(x = site, y = median_cv, fill = site)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = mean(df_cv$median_cv),
             color = "black", linetype = "dashed") +
  scale_fill_manual(values = site_colors) +
  labs(
    title    = "Within-site variability — median CV per site",
    subtitle = "Lower = daphea from same site are metabolically consistent. All sites ~0.03-0.045 = excellent reproducibility.",
    x = "Site", y = "Median coefficient of variation"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

cat("\nCV per site:\n")
print(setNames(round(within_cv, 5), sites_danube))
```

---

**Interpretation:** All sites sit between 0.030–0.045. This is outstanding biological reproducibility — individual Daphnia from the same site have nearly identical metabolic profiles. CV of 4% means the typical metabolite varies by 4% of its mean across the 12 animals at a site. Your statistical tests will be well-powered at every site.

## Heatmap — top 100 variable features

```{r heatmap, fig.height=10, fig.width=13}
top100   <- order(feat_var, decreasing = TRUE)[1:100]
heat_mat <- as.matrix(pos_mat[top100, ])

ann_col <- data.frame(
  Site  = meta[colnames(heat_mat), "site"],
  Batch = meta[colnames(heat_mat), "batch"],
  row.names = colnames(heat_mat)
)

# Build colour lists that EXACTLY match the data values
# (mismatched names cause the pheatmap error you saw before)
sites_in_data   <- unique(ann_col$Site)
batches_in_data <- unique(ann_col$Batch)
batch_pal       <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3")
batch_colors    <- setNames(batch_pal[seq_along(batches_in_data)], batches_in_data)

ann_colors <- list(
  Site  = site_colors[names(site_colors) %in% sites_in_data],
  Batch = batch_colors
)

pheatmap(
  heat_mat,
  annotation_col    = ann_col,
  annotation_colors = ann_colors,
  show_rownames     = FALSE,
  show_colnames     = FALSE,
  scale             = "row",
  clustering_method = "ward.D2",
  color = colorRampPalette(c("#313695", "#f7f7f7", "#a50026"))(100),
  main  = "Top 100 most variable features × 149 samples\n(row Z-scored | Ward's D2 clustering)"
)
```

---

**How to read this:** Each row = one metabolite. Each column = one Daphnia. **Blue = below average for that metabolite. Red = above average.** The colour is Z-scored per row so you are looking at relative patterns, not absolute amounts. If site D12 columns all go red in the same row cluster, those metabolites are specifically elevated at D12. Row clusters = groups of metabolites that move together — if those rows all map to the same KEGG pathway, that pathway is responding as a unit.

## MA plot — global Danube vs control

```{r ma-plot}
ctrl_cols   <- rownames(meta)[meta$type == "Control"]
danube_cols <- rownames(meta)[meta$type == "Danube"]

ctrl_means   <- rowMeans(pos_mat[ , ctrl_cols])
danube_means <- rowMeans(pos_mat[ , danube_cols])
fc           <- as.numeric(danube_means - ctrl_means)

df_ma <- data.frame(
  feature     = rownames(pos_mat),
  mean_ctrl   = as.numeric(ctrl_means),
  fold_change = fc
)

up   <- sum(fc >  0.5)
down <- sum(fc < -0.5)

ggplot(df_ma, aes(x = mean_ctrl, y = fold_change)) +
  geom_point(alpha = 0.3, size = 0.8, color = "#555555") +
  geom_hline(yintercept =  0,   color = "black", linewidth = 0.5) +
  geom_hline(yintercept =  0.5, color = "red",  linetype = "dashed", linewidth = 0.6) +
  geom_hline(yintercept = -0.5, color = "blue", linetype = "dashed", linewidth = 0.6) +
  annotate("text", x = 13.5, y =  0.85,
           label = paste0("↑ ", up, " elevated"), color = "#c0392b", size = 3.5) +
  annotate("text", x = 13.5, y = -0.85,
           label = paste0("↓ ", down, " depleted"), color = "#2980b9", size = 3.5) +
  labs(
    title    = "MA plot — All Danube vs Control (pooled)",
    subtitle = "Y = log fold change. Points above red = elevated in Danube. Note: pools all 12 sites — site-specific analysis follows.",
    x = "Mean control intensity", y = "Log fold change (Danube − Control)"
  ) +
  theme_bw(base_size = 12)

cat("Elevated in Danube (FC > +0.5):", up, "\n")
cat("Depleted in Danube  (FC < -0.5):", down, "\n")
```

---

**Interpretation:** 116 elevated vs 48 depleted — 2.4× more induction than suppression. Organisms under chemical stress upregulate detoxification and antioxidant pathways (producing new metabolites) more than they shut down existing ones. The downward tilt at high mean intensity means the most abundant metabolites (core intermediates — TCA cycle, amino acids) are slightly depleted in Danube animals. These are being diverted into stress-response pathways at the cost of normal housekeeping metabolism. This is a classic metabolic trade-off under environmental stress. **Important caveat:** this pools all 12 sites. The volcano plots below show the real site-specific picture.

## Site-level profiles

```{r site-profiles}
top50_idx  <- order(feat_var, decreasing = TRUE)[1:50]
top50_mat  <- pos_mat[top50_idx, ]
site_order <- c("CK", paste0("D", sprintf("%02d", 1:12)))

site_list <- lapply(site_order, function(s) {
  cols <- rownames(meta)[meta$site == s]
  cols <- cols[cols %in% colnames(top50_mat)]
  if (length(cols) == 0) return(NULL)
  data.frame(site = s, animal = cols,
             mean_int = colMeans(top50_mat[ , cols, drop = FALSE]))
})
df_site      <- do.call(rbind, Filter(Negate(is.null), site_list))
df_site$site <- factor(df_site$site, levels = site_order)

ggplot(df_site, aes(x = site, y = mean_int, fill = site)) +
  geom_boxplot(alpha = 0.85, outlier.size = 0.8) +
  scale_fill_manual(values = site_colors) +
  labs(
    title    = "Top-50 feature intensity by site (upstream → downstream)",
    subtitle = "Each point = one sample. CK = lab control. D01 = most upstream. D12 = most downstream.",
    x = "Site", y = "Mean intensity (top 50 variable features)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")
```

---

**Key findings:** D01 sits clearly above all other sites including controls. Despite being upstream, D01 shows the highest metabolic elevation — something at D01 drives broad metabolite production. Cross-check with the chemical data: what specific contaminants are elevated at D01? D11 sits with a median similar to CK, validating it as the clean Danube reference site. No linear gradient D01→D12 — confirms the chemical data finding that Danube contamination is non-linear (patchwork of industrial inputs, not a smooth gradient).

---

# ================================================================
# PART 4 — DIFFERENTIAL ABUNDANCE ANALYSIS
# ================================================================

**What this section does:** Identifies which specific metabolite features are statistically different across sites. Two approaches: (1) ANOVA across all 13 groups simultaneously — gives a global significant feature list. (2) Per-site t-tests vs control — gives volcano plots showing exactly which features are up or down at each site.

## ANOVA — global test across all sites

```{r anova}
site_factor <- meta[colnames(pos_mat), "site"]

cat("Running ANOVA on", nrow(pos_mat), "features across",
    length(unique(site_factor)), "groups...\n")

anova_F    <- numeric(nrow(pos_mat))
anova_pval <- numeric(nrow(pos_mat))

for (i in seq_len(nrow(pos_mat))) {
  y              <- as.numeric(pos_mat[i, ])
  fit            <- aov(y ~ site_factor)
  smry           <- summary(fit)[[1]]
  anova_F[i]    <- smry["site_factor", "F value"]
  anova_pval[i] <- smry["site_factor", "Pr(>F)"]
}

# Benjamini-Hochberg FDR correction
# Controls the false discovery rate: expected proportion of your
# significant list that are false positives
anova_padj <- p.adjust(anova_pval, method = "BH")

anova_results <- data.frame(
  feature = rownames(pos_mat),
  mz      = mz_values,
  F_stat  = anova_F,
  pval    = anova_pval,
  padj    = anova_padj,
  stringsAsFactors = FALSE
)
anova_results <- anova_results[order(anova_results$padj), ]

cat("\n=== ANOVA RESULTS ===\n")
cat("Significant (FDR < 0.05):", sum(anova_padj < 0.05), "\n")
cat("Significant (FDR < 0.01):", sum(anova_padj < 0.01), "\n")
cat("Significant (FDR < 0.10):", sum(anova_padj < 0.10), "\n\n")
cat("Top 10 most significant:\n")
top10 <- head(anova_results, 10)
top10[ , c("F_stat","pval","padj")] <- round(top10[ , c("F_stat","pval","padj")], 5)
print(top10[ , c("feature","mz","F_stat","pval","padj")])
```

---

**Interpretation:** 100 features significant at FDR < 0.05. This means you expect ~5 false positives in your list of 100 — 95 are real. 7.8% of all features (100/1285) showing site-level differences is biologically reasonable given the overlapping PCA clusters. The ANOVA is stringent because it must detect differences consistent across the within-site animal-to-animal noise.

```{r anova-plot}
anova_results$sig <- anova_results$padj < 0.05

ggplot(anova_results, aes(x = F_stat, y = -log10(padj), color = sig)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_hline(yintercept = -log10(0.05),
             color = "red", linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(
    values = c("FALSE" = "#aaaaaa", "TRUE" = "#d62728"),
    labels = c("Not significant",
               paste0("FDR < 0.05 (n=", sum(anova_results$padj < 0.05), ")"))
  ) +
  labs(
    title    = "ANOVA — 1285 features across 13 site groups",
    subtitle = "X = F-statistic (larger = bigger between-site signal vs within-site noise). Y = significance.",
    x = "F-statistic", y = "-log10(FDR adjusted p-value)", color = ""
  ) +
  theme_bw(base_size = 12)
```

## PERMANOVA — whole-metabolome test

```{r permanova}
# ANOVA tests features one-by-one. PERMANOVA tests ALL 1285 simultaneously:
# "Does the overall metabolite composition differ by site?"
# This is the multivariate headline test for your research question.

mat_t <- t(pos_mat)   # vegan needs rows=samples, cols=features

set.seed(42)
perm_result <- adonis2(
  mat_t ~ site_factor,
  method       = "euclidean",
  permutations = 999
)

cat("=== PERMANOVA ===\n")
print(perm_result)
cat("\nR² =", round(perm_result$R2[1], 4),
    "→ site explains", round(100 * perm_result$R2[1], 1),
    "% of total metabolome variance\n")
```

---

**Interpretation:** R² tells you effect size. If R² ≈ 0.15–0.25 and p < 0.05, site explains 15–25% of total metabolic variation — moderate and realistic for a field study. Write in your report: *"PERMANOVA confirmed that site identity significantly explained metabolome composition (R²=X%, p<0.001, 999 permutations), consistent with site-specific chemical mixture exposures producing distinct metabolic signatures."*

## Volcano plots — each site vs control

```{r volcano}
run_volcano <- function(site_name) {
  ck_cols   <- rownames(meta)[meta$site == "CK"]
  site_cols <- rownames(meta)[meta$site == site_name]
  site_cols <- site_cols[site_cols %in% colnames(pos_mat)]

  fc   <- rowMeans(pos_mat[ , site_cols]) - rowMeans(pos_mat[ , ck_cols])
  pval <- sapply(seq_len(nrow(pos_mat)), function(i) {
    tryCatch(
      t.test(as.numeric(pos_mat[i, site_cols]),
             as.numeric(pos_mat[i, ck_cols]))$p.value,
      error = function(e) NA_real_
    )
  })

  padj <- p.adjust(pval, method = "BH")
  cat_label <- ifelse(padj < 0.05 & fc >  0.5, "Up in Danube",
               ifelse(padj < 0.05 & fc < -0.5, "Down in Danube",
                                                "Not significant"))

  data.frame(feature = rownames(pos_mat), site = site_name,
             fold_change = fc, pval = pval, padj = padj,
             neg_logp = -log10(padj), category = cat_label,
             stringsAsFactors = FALSE)
}

cat("Running per-site comparisons vs control...\n")
all_volcano <- do.call(rbind, lapply(
  paste0("D", sprintf("%02d", 1:12)), run_volcano
))
cat("Done.\n\n")

# Summary table
cat("=== SIGNIFICANT FEATURES PER SITE (FDR<0.05, |FC|>0.5) ===\n")
vsummary <- do.call(rbind, lapply(paste0("D", sprintf("%02d", 1:12)), function(s) {
  sub <- all_volcano[all_volcano$site == s, ]
  data.frame(Site  = s,
             Total = sum(sub$category != "Not significant"),
             Up    = sum(sub$category == "Up in Danube"),
             Down  = sum(sub$category == "Down in Danube"))
}))
print(vsummary)
```

```{r volcano-grid, fig.height=12, fig.width=14}
ggplot(all_volcano, aes(x = fold_change, y = neg_logp, color = category)) +
  geom_point(alpha = 0.4, size = 0.7) +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = c(-0.5, 0.5),
             linetype = "dashed", color = "grey50", linewidth = 0.4) +
  scale_color_manual(
    values = c("Not significant" = "#cccccc",
               "Up in Danube"   = "#d62728",
               "Down in Danube" = "#2c7bb6")
  ) +
  facet_wrap(~ site, ncol = 4) +
  labs(
    title    = "Volcano plots — each Danube site vs Control",
    subtitle = "X = fold change (positive = higher than control). Y = significance. Coloured = FDR<0.05 and |FC|>0.5.",
    x = "Log fold change (site − control)",
    y = "-log10(FDR adjusted p-value)",
    color = ""
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 11))
```

---

**What the volcano summary reveals:**

- **D12: 147 significant** — the most complex and strongest metabolic response. Most downstream site, most chemicals.
- **D09: 126 sig (105 up, 21 down)** — dominant upregulation. Strong stress-response activation.
- **D04: 113 sig** — strong mixed response (75 up, 38 down).
- **D02: 105 sig** — strong induction-dominant response.
- **D05, D06, D07, D08: 0 significant** — metabolically indistinguishable from controls. Their chemical mixtures either have limited metabolic detectability or the animal-to-animal variation swamps any site signal.
- **D01: only 2 significant (both down)** — paradox. D01 had the *highest* mean intensity in site profiles but almost no specific significant features. Reason: D01's response is a broad, non-specific elevation across many features (all slightly up), not a few features strongly up. A per-feature t-test requires specific large changes, not broad small ones. D01's signal is in the *distribution*, not in individual features.

## Heatmap of significant features only

```{r sig-heatmap, fig.height=9, fig.width=13}
sig_features <- anova_results$feature[anova_results$padj < 0.05]
sig_mat      <- pos_mat[sig_features, ]

cat("Significant features:", length(sig_features), "\n")

heat_sig <- as.matrix(sig_mat)

ann_col2 <- data.frame(
  Site  = meta[colnames(heat_sig), "site"],
  Batch = meta[colnames(heat_sig), "batch"],
  row.names = colnames(heat_sig)
)

sites2   <- unique(ann_col2$Site)
batches2 <- unique(ann_col2$Batch)
b_colors2 <- setNames(batch_pal[seq_along(batches2)], batches2)

ann_colors2 <- list(
  Site  = site_colors[names(site_colors) %in% sites2],
  Batch = b_colors2
)

pheatmap(
  heat_sig,
  annotation_col    = ann_col2,
  annotation_colors = ann_colors2,
  show_rownames     = FALSE,
  show_colnames     = FALSE,
  scale             = "row",
  clustering_method = "ward.D2",
  color = colorRampPalette(c("#313695", "#f7f7f7", "#a50026"))(100),
  main  = paste0("Statistically significant features only (n=", length(sig_features),
                 ", FDR<0.05)\nRow Z-scored | Ward's D2 clustering")
)
```

---

**Why this heatmap is more meaningful than the EDA one:** The EDA heatmap showed the top 100 by variance — variance alone doesn't mean significance. A feature could be variable because 3 animals happened to be outliers. This heatmap shows only the 100 features that passed FDR<0.05 — every row has statistical evidence of site-level difference. Row clusters here represent genuinely co-regulated metabolite groups. When you annotate those clusters with KEGG pathway IDs, those are your biological findings.

---

# ================================================================
# PART 5 — PATHWAY ANNOTATION & TEAM HANDOFF
# ================================================================

## Join significant features to annotations

```{r pathway-join}
# Merge annotation tables
annot_full <- merge(
  pos_kegg[ , c("feature_id", "kegg_ids")],
  pos_hmdb[ , c("feature_id", "hmdb_ids")],
  by = "feature_id", all = TRUE
)

# Join to ANOVA results
sig_annotated <- merge(
  anova_results[anova_results$padj < 0.05, ],
  annot_full,
  by.x = "feature", by.y = "feature_id",
  all.x = TRUE
)

cat("Significant features (FDR<0.05):      ", nrow(sig_annotated), "\n")
cat("Of which KEGG annotated:              ",
    sum(!is.na(sig_annotated$kegg_ids)), "\n")
cat("Of which HMDB annotated:              ",
    sum(!is.na(sig_annotated$hmdb_ids)), "\n")
cat("Of which unannotated (used in ML):    ",
    sum(is.na(sig_annotated$kegg_ids) & is.na(sig_annotated$hmdb_ids)), "\n\n")

cat("Top 15 significant features with KEGG annotation:\n")
top_kegg <- sig_annotated[!is.na(sig_annotated$kegg_ids), ]
top_kegg <- top_kegg[order(top_kegg$padj), ]
show <- top_kegg[1:min(15, nrow(top_kegg)),
                 c("feature", "mz", "F_stat", "padj", "kegg_ids", "hmdb_ids")]
show[ , c("F_stat","padj")] <- round(show[ , c("F_stat","padj")], 5)
print(show)
```

```{r kegg-frequency}
# Which KEGG compound IDs appear most in your significant features?
# High frequency = implicated by multiple features = likely important

sig_kegg_ids <- unlist(strsplit(
  sig_annotated$kegg_ids[!is.na(sig_annotated$kegg_ids)], ";"))
sig_kegg_ids <- trimws(sig_kegg_ids[nchar(trimws(sig_kegg_ids)) > 0])

kegg_freq <- sort(table(sig_kegg_ids), decreasing = TRUE)

cat("Most frequent KEGG compound IDs in significant features:\n")
print(head(kegg_freq, 20))

# Look these up at: https://www.genome.jp/kegg/compound/
# e.g., search "C00158" → citric acid → TCA cycle
# Ecotox pathways to watch: steroid biosynthesis, glutathione,
# TCA cycle, amino acid metabolism, fatty acid metabolism

df_kegg_freq <- data.frame(
  kegg_id = names(kegg_freq),
  count   = as.integer(kegg_freq)
)

ggplot(head(df_kegg_freq, 20), aes(x = reorder(kegg_id, count), y = count)) +
  geom_col(fill = "#2c7bb6", alpha = 0.85) +
  coord_flip() +
  labs(
    title    = "Most frequent KEGG IDs among significant features",
    subtitle = "Look these up at kegg.jp/compound to get metabolite names and pathways.",
    x = "KEGG Compound ID", y = "Feature count"
  ) +
  theme_bw(base_size = 11)
```

```{r}
# ================================================================
# ANNOTATION MATCHING — CORRECTED
# Uses your own R ANOVA results, NOT MetaboAnalyst output
# MetaboAnalyst renames features internally so IDs don't match
# ================================================================

setwd("C:/Users/hrush/Desktop/UoB/LECTURE/Computational Biology/Data Jamboree")

# Load annotation files
kegg_annot <- read.table(
  "polar_pos_pkl_to_kegg_annotations.tsv",
  sep = "\t", header = FALSE,
  col.names = c("feature_id", "mz", "kegg_ids"),
  stringsAsFactors = FALSE,
  na.strings = c("", "NA")
)

hmdb_annot <- read.table(
  "polar_pos_pkl_to_hmdb_annotations.tsv",
  sep = "\t", header = FALSE,
  col.names = c("feature_id", "mz", "hmdb_ids"),
  stringsAsFactors = FALSE,
  na.strings = c("", "NA")
)

cat("Total features in annotation file:", nrow(kegg_annot), "\n")
cat("KEGG annotated:", sum(!is.na(kegg_annot$kegg_ids)), "\n")
cat("HMDB annotated:", sum(!is.na(hmdb_annot$hmdb_ids)), "\n\n")

# ---- Use your R ANOVA results (anova_results is already in workspace) ----
# anova_results was built in Part 4 of your Rmd
# It has columns: feature, mz, F_stat, pval, padj
# feature column matches annotation file IDs exactly

sig_features_r <- anova_results[anova_results$padj < 0.05, ]
cat("Significant features from R ANOVA (site-based):", nrow(sig_features_r), "\n\n")

# Merge KEGG annotations
sig_kegg <- merge(
  sig_features_r,
  kegg_annot[ , c("feature_id", "kegg_ids")],
  by.x = "feature", by.y = "feature_id",
  all.x = TRUE
)

# Merge HMDB annotations
sig_annot <- merge(
  sig_kegg,
  hmdb_annot[ , c("feature_id", "hmdb_ids")],
  by.x = "feature", by.y = "feature_id",
  all.x = TRUE
)

cat("=== ANNOTATION RESULTS ===\n")
cat("Total significant:", nrow(sig_annot), "\n")
cat("With KEGG ID:     ", sum(!is.na(sig_annot$kegg_ids)), "\n")
cat("With HMDB ID:     ", sum(!is.na(sig_annot$hmdb_ids)), "\n")
cat("Unannotated:      ", sum(is.na(sig_annot$kegg_ids) & 
                               is.na(sig_annot$hmdb_ids)), "\n\n")

# Show top annotated features
top_kegg <- sig_annot[!is.na(sig_annot$kegg_ids), ]
top_kegg <- top_kegg[order(top_kegg$padj), ]
cat("Top significant features with KEGG annotation:\n")
print(top_kegg[ , c("feature", "mz", "F_stat", "padj", "kegg_ids")])

# ---- Extract KEGG IDs for MetaboAnalyst pathway analysis ----

# Split semicolon-separated IDs and flatten
kegg_split <- unlist(strsplit(
  top_kegg$kegg_ids[!is.na(top_kegg$kegg_ids)], ";"))
kegg_split <- trimws(kegg_split)
kegg_split <- kegg_split[nchar(kegg_split) > 0]
kegg_unique <- unique(kegg_split)

cat("\nUnique KEGG IDs for pathway analysis:", length(kegg_unique), "\n")
cat("IDs:\n")
cat(paste(kegg_unique, collapse = "\n"))

# Save KEGG list as plain text
writeLines(as.character(kegg_unique), "sig_kegg_ids_site.txt")
cat("\n\nSaved: sig_kegg_ids_site.txt\n")

# Save full annotated table
write.csv(sig_annot, "sig_features_site_annotated.csv", row.names = FALSE)
cat("Saved: sig_features_site_annotated.csv\n")
```

```{r}
# FOR ANNOTATION MATCHING — POSITIVE MODE (MetaboAnalyst anova_posthoc.csv)
setwd("C:/Users/hrush/Desktop/UoB/LECTURE/Computational Biology/Data Jamboree")

library(dplyr)
library(tidyr)

# Load ANOVA results from MetaboAnalyst
anova_res <- read.csv("anova_posthoc.csv", row.names = 1)
sig <- anova_res[anova_res$FDR < 0.05, ]
cat("Significant features:", nrow(sig), "\n")

# Extract mz values from MetaboAnalyst feature IDs
# MetaboAnalyst names them "mz_326.0856" — strip the "mz_" prefix
sig$mz_val <- as.numeric(sub("mz_", "", rownames(sig)))
sig$feature_id <- rownames(sig)

# Load annotation files
kegg_annot <- read.delim(
  "polar_pos_pkl_to_kegg_annotations.tsv",
  header = FALSE, sep = "\t",
  col.names = c("feature_id_annot", "mz_annot", "kegg_ids"),
  stringsAsFactors = FALSE,
  fill = TRUE, quote = "",
  na.strings = c("", "NA")
)

hmdb_annot <- read.delim(
  "polar_pos_pkl_to_hmdb_annotations.tsv",
  header = FALSE, sep = "\t",
  col.names = c("feature_id_annot", "mz_annot", "hmdb_ids"),
  stringsAsFactors = FALSE,
  fill = TRUE, quote = "",
  na.strings = c("", "NA")
)

cat("Total pos features:", nrow(kegg_annot), "\n")
cat("Features with KEGG IDs:", sum(!is.na(kegg_annot$kegg_ids)), "\n")
cat("Features with HMDB IDs:", sum(!is.na(hmdb_annot$hmdb_ids)), "\n\n")

# ---- KEY FIX: match on mz value, not feature ID string ----
# MetaboAnalyst uses "mz_326.0856", annotation uses "326_0856092"
# These encode the same mz but as different strings
# Solution: match numerically within tolerance 0.001 Da

tolerance <- 0.001

match_by_mz <- function(sig_df, annot_df, annot_col) {
  result_col <- rep(NA_character_, nrow(sig_df))
  for (i in seq_len(nrow(sig_df))) {
    diffs <- abs(annot_df$mz_annot - sig_df$mz_val[i])
    best  <- which.min(diffs)
    if (diffs[best] < tolerance) {
      result_col[i] <- annot_df[[annot_col]][best]
    }
  }
  result_col
}

sig$kegg_ids <- match_by_mz(sig, kegg_annot, "kegg_ids")
sig$hmdb_ids <- match_by_mz(sig, hmdb_annot, "hmdb_ids")

# Results
cat("Significant with KEGG ID:", sum(!is.na(sig$kegg_ids)), "\n")
cat("Significant with HMDB ID:", sum(!is.na(sig$hmdb_ids)), "\n")
cat("Unannotated:             ", 
    sum(is.na(sig$kegg_ids) & is.na(sig$hmdb_ids)), "\n\n")

cat("Annotated KEGG %:", 
    round(100 * sum(!is.na(sig$kegg_ids)) / nrow(sig), 1), "%\n")
cat("Annotated HMDB %:", 
    round(100 * sum(!is.na(sig$hmdb_ids)) / nrow(sig), 1), "%\n\n")

# Show annotated features
cat("Features with KEGG annotation:\n")
kegg_hits <- sig[!is.na(sig$kegg_ids), c("mz_val", "p.value", "FDR", "kegg_ids", "hmdb_ids")]
kegg_hits <- kegg_hits[order(kegg_hits$FDR), ]
print(kegg_hits)

# Save annotated table
write.csv(sig, "sig_features_pos_annotated.csv", row.names = TRUE)
cat("\nSaved: sig_features_pos_annotated.csv\n")

# Extract KEGG IDs for MetaboAnalyst pathway analysis
sig_kegg_expanded <- sig %>%
  filter(!is.na(kegg_ids) & kegg_ids != "") %>%
  separate_rows(kegg_ids, sep = ";") %>%
  filter(trimws(kegg_ids) != "")

kegg_list <- unique(trimws(sig_kegg_expanded$kegg_ids))
kegg_list <- as.character(kegg_list[nchar(kegg_list) > 0])

writeLines(kegg_list, "pos_sig_kegg_ids.txt")
cat("KEGG IDs saved to: pos_sig_kegg_ids.txt\n")
cat("Total unique KEGG IDs:", length(kegg_list), "\n")
cat("\nPaste these into MetaboAnalyst → Pathway Analysis:\n")
cat(paste(kegg_list, collapse = "\n"))
```

```{r}
# FOR ANNOTATION MATCHING — POSITIVE MODE (MetaboAnalyst anova_posthoc.csv)
setwd("C:/Users/hrush/Desktop/UoB/LECTURE/Computational Biology/Data Jamboree")
library(dplyr)
library(tidyr)

# Load ANOVA results from MetaboAnalyst
anova_res <- read.csv("anova_posthoc.csv", row.names = 1)
sig <- anova_res[anova_res$FDR < 0.05, ]
cat("Significant features:", nrow(sig), "\n")

# Extract mz values from MetaboAnalyst feature IDs
sig$mz_val <- as.numeric(sub("mz_", "", rownames(sig)))
sig$feature_id <- rownames(sig)

# Load annotation files
kegg_annot <- read.delim(
  "polar_pos_pkl_to_kegg_annotations.tsv",
  header = FALSE, sep = "\t",
  col.names = c("feature_id_annot", "mz_annot", "kegg_ids"),
  stringsAsFactors = FALSE,
  fill = TRUE, quote = "",
  na.strings = c("", "NA")
)
hmdb_annot <- read.delim(
  "polar_pos_pkl_to_hmdb_annotations.tsv",
  header = FALSE, sep = "\t",
  col.names = c("feature_id_annot", "mz_annot", "hmdb_ids"),
  stringsAsFactors = FALSE,
  fill = TRUE, quote = "",
  na.strings = c("", "NA")
)

cat("Total pos features:", nrow(kegg_annot), "\n")
cat("Features with KEGG IDs:", sum(!is.na(kegg_annot$kegg_ids)), "\n")
cat("Features with HMDB IDs:", sum(!is.na(hmdb_annot$hmdb_ids)), "\n\n")

# Match on mz value within tolerance
tolerance <- 0.001

match_by_mz <- function(sig_df, annot_df, annot_col) {
  result_col <- rep(NA_character_, nrow(sig_df))
  for (i in seq_len(nrow(sig_df))) {
    diffs <- abs(annot_df$mz_annot - sig_df$mz_val[i])
    best  <- which.min(diffs)
    if (diffs[best] < tolerance) {
      result_col[i] <- annot_df[[annot_col]][best]
    }
  }
  result_col
}

sig$kegg_ids <- match_by_mz(sig, kegg_annot, "kegg_ids")
sig$hmdb_ids <- match_by_mz(sig, hmdb_annot, "hmdb_ids")

# Summary
cat("Significant with KEGG ID:", sum(!is.na(sig$kegg_ids)), "\n")
cat("Significant with HMDB ID:", sum(!is.na(sig$hmdb_ids)), "\n")
cat("Unannotated:             ", 
    sum(is.na(sig$kegg_ids) & is.na(sig$hmdb_ids)), "\n\n")
cat("Annotated KEGG %:", 
    round(100 * sum(!is.na(sig$kegg_ids)) / nrow(sig), 1), "%\n")
cat("Annotated HMDB %:", 
    round(100 * sum(!is.na(sig$hmdb_ids)) / nrow(sig), 1), "%\n\n")

# Show annotated features
cat("Features with KEGG annotation:\n")
kegg_hits <- sig[!is.na(sig$kegg_ids), c("mz_val", "p.value", "FDR", "kegg_ids", "hmdb_ids")]
kegg_hits <- kegg_hits[order(kegg_hits$FDR), ]
print(kegg_hits)

# Save annotated table
write.csv(sig, "sig_features_pos_annotated.csv", row.names = TRUE)
cat("\nSaved: sig_features_pos_annotated.csv\n")

# ---- KEGG IDs for pathway analysis ----
sig_kegg_expanded <- sig %>%
  filter(!is.na(kegg_ids) & kegg_ids != "") %>%
  separate_rows(kegg_ids, sep = ";") %>%
  filter(trimws(kegg_ids) != "")

kegg_list <- unique(trimws(sig_kegg_expanded$kegg_ids))
kegg_list <- as.character(kegg_list[nchar(kegg_list) > 0])
writeLines(kegg_list, "pos_sig_kegg_ids.txt")
cat("KEGG IDs saved to: pos_sig_kegg_ids.txt\n")
cat("Total unique KEGG IDs:", length(kegg_list), "\n\n")

# ================================================================
# NEW — HMDB IDs for MetaboAnalyst MSEA
# ================================================================

# Extract and clean all HMDB IDs from significant features
sig_hmdb_expanded <- sig %>%
  filter(!is.na(hmdb_ids) & hmdb_ids != "") %>%
  separate_rows(hmdb_ids, sep = ";") %>%
  mutate(hmdb_ids = trimws(hmdb_ids)) %>%
  filter(nchar(hmdb_ids) > 0)

hmdb_list <- unique(sig_hmdb_expanded$hmdb_ids)
hmdb_list <- as.character(hmdb_list[nchar(hmdb_list) > 0])

cat("=== HMDB IDs for MetaboAnalyst MSEA ===\n")
cat("Total unique HMDB IDs:", length(hmdb_list), "\n\n")

# MetaboAnalyst MSEA accepts HMDB IDs in format HMDB0000001
# Your IDs are already in this format — no conversion needed
cat("Your HMDB IDs (paste into MetaboAnalyst MSEA):\n")
cat(paste(hmdb_list, collapse = "\n"))

# Save to file
writeLines(hmdb_list, "pos_sig_hmdb_ids.txt")
cat("\n\nSaved: pos_sig_hmdb_ids.txt\n")

# Also extract background HMDB IDs (all annotated features)
# MetaboAnalyst needs this for a fair enrichment test
bg_hmdb_expanded <- hmdb_annot %>%
  filter(!is.na(hmdb_ids) & hmdb_ids != "") %>%
  separate_rows(hmdb_ids, sep = ";") %>%
  mutate(hmdb_ids = trimws(hmdb_ids)) %>%
  filter(nchar(hmdb_ids) > 0)

bg_hmdb_list <- unique(as.character(bg_hmdb_expanded$hmdb_ids))
writeLines(bg_hmdb_list, "pos_background_hmdb_ids.txt")
cat("Background HMDB IDs saved to: pos_background_hmdb_ids.txt\n")
cat("Total background HMDB IDs:", length(bg_hmdb_list), "\n")

# Summary table of all annotated significant features
cat("\n=== FULL ANNOTATED SIGNIFICANT FEATURES ===\n")
annotated_sig <- sig[!is.na(sig$kegg_ids) | !is.na(sig$hmdb_ids), 
                      c("mz_val", "p.value", "FDR", "kegg_ids", "hmdb_ids")]
annotated_sig <- annotated_sig[order(annotated_sig$FDR), ]
print(annotated_sig)
```

## Save all outputs

```{r save-all}
# Save everything your team needs
sig_mat_shared <- sig_mat[ , !colnames(sig_mat) %in%
                              c("D06A2","D06B3","D07A3","D08B1","D10B4")]

save(
  pos_mat, mz_values, meta, pos_kegg, pos_hmdb, site_colors,
  pca_df, pca_var, pca_res,
  feat_var, feat_mean,
  anova_results, sig_features, sig_mat, sig_annotated,
  all_volcano, vsummary,
  sig_mat_shared,
  file = "metabolomics_complete.RData"
)

write.csv(sig_annotated,
          "significant_features_annotated_pos.csv",
          row.names = FALSE)
write.csv(vsummary,
          "site_significant_counts_pos.csv",
          row.names = FALSE)

cat("Saved files:\n")
cat("  metabolomics_complete.RData           (full R workspace)\n")
cat("  significant_features_annotated_pos.csv (100 sig features + annotations)\n")
cat("  site_significant_counts_pos.csv        (per-site summary for team)\n")
```

## Messages for your teammates

```{r team-handoff, echo=FALSE}
cat("=== FOR YOUR PARTNER (negative mode) ===\n")
cat("Shared samples (for combining modes): 144\n")
cat("Samples in your positive only: D06A2, D06B3, D07A3, D08B1, D10B4\n")
cat("Sample in negative only: D01B4\n")
cat("When combining: shared <- intersect(colnames(pos_mat), colnames(neg_mat))\n\n")

cat("=== FOR TRANSCRIPTOMICS ===\n")
cat("Our PCA: sites overlap, no clean separation.\n")
cat("Strongest metabolic signals: D02, D04, D09, D10, D12.\n")
cat("Compare with RNA-seq PCA — do the same sites drive variation?\n")
cat("D01 paradox: highest mean elevation but 0 specific significant features.\n\n")

cat("=== FOR CHEMICAL ANALYSIS ===\n")
cat("D01: broad metabolic elevation, no specific features — what's unique at D01 chemically?\n")
cat("D05/D06/D07/D08: 0 significant features — what do these mixtures share?\n")
cat("D12: most complex response (147 sig features) — most complex mixture?\n\n")

cat("=== FOR ML / INTEGRATION ===\n")
cat("Deliverable 1: sig_mat_shared (100 features × 144 shared samples)\n")
cat("Deliverable 2: significant_features_annotated_pos.csv\n")
cat("Deliverable 3: site_significant_counts_pos.csv\n")
cat("Suggestion: first 5 PCA scores as dimensionality-reduced input for iRF\n")
cat("Use all 1285 features (not just significant 100) for ML — let iRF select.\n")
```

## Final summary table

```{r final-summary, echo=FALSE}
summary_df <- data.frame(
  Item = c(
    "Features", "Samples", "Missing values", "Near-zero variance features",
    "KEGG annotated", "HMDB annotated", "Unannotated",
    "PCA outliers", "QC flags (Layer 1 mean)",
    "ANOVA significant (FDR<0.05)", "ANOVA significant (FDR<0.01)",
    "Sites with 0 sig features", "Site with most sig features",
    "Samples for multi-omics integration"
  ),
  Result = c(
    "1,285", "149 (positive mode)", "0 — fully imputed",
    "0 — all retained", "437 (34.0%)", "409 (31.8%)", "757 (58.9%)",
    "1 — D08B6 (biological, kept)", "5 — marginal, all kept",
    "100", "45",
    "D05, D06, D07, D08",
    "D12 (147 sig, 103 up, 44 down)",
    "144 shared with negative mode"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(summary_df, caption = "Complete Analysis Summary — Positive Mode Metabolomics")
```

---

## Session info

```{r session-info}
sessionInfo()
```
